### 2021-11-23 draft.tex
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@b40d93aa6c51b5048d941a37d5ef873e2adb69a2 🚀
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@765219f986f193eae84c6d52a51418c5a37818c4 🚀
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@5099287b68fe73e39b47b3d4744b3c0aadd2b4aa 🚀

Within the last several hundred years, human activity has induced rapid
changes in Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrain and oceans, and both agricultural and urban development has
rapidly reshaped the Earth's land cover. These the bulk of this change
has occurred within the last several hundred years, a geological
instant, inducing a sudden shift in conditions to Earth's climate and
biosphere. As a result, predicting how ecosystems will change in the
future, \emph{ecological forecasting}, and then using these forecasts to
make decisions to mitigate the negative consequences of this change on
ecosystems, their functioning, and the services they provide to humans
has emerged as an imperative for ecology and environmental science
(Dietze 2017). However, robust prediction of ecological processes is, to
say the least, quite difficult (Beckage \emph{et al.} 2011; Petchey
\emph{et al.} 2015). This difficultly is compounded by a few factors,
the first being that sampling ecosystems is not easy. Ecological data is
often biased, noisey, and sparse in both space and time. The current
paucity of ecological data has resulted in much interest in developing
global systems for \emph{ecosystem monitoring} (Makiola \emph{et al.}
2020), which would systematize the collection of biodiversity data in
manner that makes detecting and predicting change more possible than at
the moment (Urban \emph{et al.} 2021).

The second major challenge in ecological forecasting is that the
underlying dynamics of most ecological processes are unknown and instead
must be inferred from this (sparse) data. Much of the history of
quantitatively modeling ecosystems have been done in the language of
dynamical systems, describing how the value of an observable state of
the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time, yielding
models in the form of differential equations in continuous-time
settings, \(\frac{dx}{dt} = f(x)\), or difference equations in
discrete-time settings, \(x_t = f(x_{t-1})\), where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Beckage \emph{et al.}
2011; Chen \emph{et al.} 2019; Pennekamp \emph{et al.} 2019). The
initial success of these forms of models can be traced back to the
larger program of ontological reductionism, which became the default
approach to modeling in the sciences after its early success in physics,
which, by the time ecology was becoming a quantitative science (sometime
in the 20th century, depending on who you ask), became the foundation
for mathematical models in ecology.

@@

\hypertarget{chapter-one-forecasting-the-spatial-uncoupling-of-a-plant-pollinator-network}{%
\section{Chapter One: Forecasting the spatial uncoupling of a
plant-pollinator
network}\label{chapter-one-forecasting-the-spatial-uncoupling-of-a-plant-pollinator-network}}

@@

\hypertarget{data}{%
\subsection{Data}\label{data}}

@@

\hypertarget{preliminary-results}{%
\subsection{Preliminary Results}\label{preliminary-results}}

@@

\hypertarget{chapter-two-optimizing-spatial-sampling-of-species-interactions}{%
\section{Chapter Two: Optimizing spatial sampling of species
interactions}\label{chapter-two-optimizing-spatial-sampling-of-species-interactions}}

This chapter uses simulation models to investigate the relationship
between species relative abundance, sampling effort, and probability of
observing an interaction between species, and further proposes a method
for optimizing the spatial sampling locations to maximize the
probability of detecting an interaction between two species given their
distributions. This addresses the optimization of monitoring network
part of the flow from data to mitigation at the top of
fig.~\ref{fig:thesis}, left. As explored in the previous chapter, there
are false-negatives in interaction data. However, there is more than one
way to observe a false-negative when sampling interactions.
fig.~\ref{fig:fnrtaxonomy} shows a taxonomy of false-negatives in
occurrence, co-occurrence, and interaction data.

Co-occurrence is not the same thing as interaction (Blanchet \emph{et
al.} 2020), but often is used as a proxy.

The first result is to compute a null expectation of the probability of
an interaction false-negative as a function number of total observations
of individuals of \emph{any species}. This is done by using a log-normal
distribution of relative abundance (Hubbell 2001) and simulating the
process of observation on food-webs generated using the niche model
(Williams \& Martinez 2000) with connectance parameterized by the
flexible-links model (MacDonald \emph{et al.} 2020). An example of this
relation for networks with varying spceies richness is shown in
fig.~\ref{fig:fnr}.

@@

We propose various landscape modifications which alter the cover of a
landscape, represented as a raster. We then compute a new resistance
surface based on the proposed landscape modification using Circuitscape
(McRae \emph{et al.} 2008), and based on the values of resistance to
dispersal between pair of locations we simulate spatially-explicit
metapopulation dynamics model (Hanski \& Ovaskainen 2000; Ovaskainen
\emph{et al.} 2002) to estimate a distribution of time until extinction
for each landscape modification. The largest challenge in implementing
this algorithm is the space of potential modifications grows as
\(O(nm)\) for an \(n\) by \(m\) raster. For most actual landscapes to
which we wish to apply this method, the set of possible modifications
becomes uncomputably large, so we use simulated annealing to explore the
search space of possible modifications to estimate the modification that
maximizes the time-until extinction of simulated metapopulation dynamics
under that hypothetical modified landscape.

@@

@@

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

\leavevmode\hypertarget{ref-Blanchet2020CooNot}{}%
Blanchet, F.G., Cazelles, K. \& Gravel, D. (2020). Co-occurrence is not
evidence of ecological interactions. \emph{Ecology Letters}, 23,
1050--1063.

>>>>

Within the last several hundred years, human activity has induced rapid
changes in Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrain and oceans, and both agricultural and urban development has
rapidly reshaped the Earth's land cover. These the bulk of this change
has occurred within the last several hundred years, a geological
instant, inducing a sudden shift in conditions to Earth's climate and
biosphere. As a result \emph{ecological forecasting}---TK DEFINITION
---and then using these forecasts to make decisions to mitigate the
negative consequences of this change on ecosystems, their functioning,
and the services they provide to humans has emerged as an imperative for
ecology and environmental science (Dietze 2017). However, robust
prediction of ecological processes is, to say the least, quite difficult
(Beckage \emph{et al.} 2011; Petchey \emph{et al.} 2015). This
difficultly is compounded by a few factors, the first being that
sampling ecosystems is not easy. Ecological data is often biased,
noisey, and sparse in both space and time. The current paucity of
ecological data has resulted in much interest in developing global
systems for \emph{ecosystem monitoring} (Makiola \emph{et al.} 2020),
which would systematize the collection of biodiversity data in manner
that makes detecting and predicting change more possible than at the
moment (Urban \emph{et al.} 2021).

\textbf{\emph{TK add mention of interaction prediction here as they
effect dynamics}} The second major challenge in ecological forecasting
is that the underlying dynamics of most ecological processes are unknown
and instead must be inferred from this (sparse) data. Much of the
history of quantitatively modeling ecosystems have been done in the
language of dynamical systems, describing how the value of an observable
state of the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time, yielding
models in the form of differential equations in continuous-time
settings, \(\frac{dx}{dt} = f(x)\), or difference equations in
discrete-time settings, \(x_t = f(x_{t-1})\), where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Beckage \emph{et al.}
2011; Chen \emph{et al.} 2019; Pennekamp \emph{et al.} 2019). The
initial success of these forms of models can be traced back to the
larger program of ontological reductionism, which became the default
approach to modeling in the sciences after its early success in physics,
which, by the time ecology was becoming a quantitative science (sometime
in the 20th century, depending on who you ask), became the foundation
for mathematical models in ecology.

@@

\textbf{TK final paragraph} \textgreater{} Set up the key research
challenges in this field so that the reader understands where your
thesis fits in. \textgreater{} Finish the introduction with a short
summary of what you will do across your chapters. This is the bridging
paragraph to your chapters.

The primary reserach challenges this thesis addresses: how do we design
ecological samples? Data assimilation pipeline. How do we propgate
uncertainty from these samples into forecasts? More details on pipeline
flow

\hypertarget{chapter-one-optimizing-spatial-sampling-of-species-interactions}{%
\section{Chapter One: Optimizing spatial sampling of species
interactions}\label{chapter-one-optimizing-spatial-sampling-of-species-interactions}}

\hypertarget{objective}{%
\subsection{Objective}\label{objective}}

This chapter uses simulation models to investigate the relationship
between species relative abundance, sampling effort, and probability of
observing an interaction between species in order to aid in the design
of samples of ecological interactions, and to provide a null expectation
of false-negative probability for a dataset of a given size. Further it
then proposes a method for optimizing the spatial sampling locations to
maximize the probability of detecting an interaction between two species
given a fixed number of total of observations, and the distributions of
each species. This addresses the optimization of monitoring network part
of the flow from data to mitigation at the top of fig.~\ref{fig:thesis},
left. As explored in the previous chapter, there are false-negatives in
interaction data. However, there is more than one way to observe a
false-negative when sampling interactions. fig.~\ref{fig:fnrtaxonomy}
shows a taxonomy of false-negatives in occurrence, co-occurrence, and
interaction data.

The first result is to compute a null expectation of the probability of
an interaction false-negative as a function number of total observations
of individuals of \emph{any species}. This is done by simulating the
process of observation, where the probability of observing a given
species is its relative abundance. We use a log-normal distribution of
relative abundance (Hubbell 2001) and simulating the process of
observation on food-webs generated using the niche model (Williams \&
Martinez 2000) with connectance parameterized by the flexible-links
model (MacDonald \emph{et al.} 2020).

An example of this relation for networks with varying spceies richness
is shown in fig.~\ref{fig:fnr}.

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\hypertarget{progress}{%
\subsection{Progress}\label{progress}}

This chapter is mostly complete. The only remaining work is the
implementation of simulated annealing optimization process.

\hypertarget{chapter-two-forecasting-the-spatial-uncoupling-of-a-plant-pollinator-network}{%
\section{Chapter Two: Forecasting the spatial uncoupling of a
plant-pollinator
network}\label{chapter-two-forecasting-the-spatial-uncoupling-of-a-plant-pollinator-network}}

@@

\hypertarget{methods-1}{%
\subsection{Methods}\label{methods-1}}

@@

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

@@

\hypertarget{progress-1}{%
\subsection{Progress}\label{progress-1}}

I have a phylogeny and SDMs for all the species. Waiting on data
agreement with RMBL.

@@

We propose various landscape modifications which alter the cover of a
landscape, represented as a raster. We then compute a new resistance
surface based on the proposed landscape modification using Circuitscape
(McRae \emph{et al.} 2008), and based on the values of resistance to
dispersal between pair of locations we simulate spatially-explicit
metapopulation dynamics model (Hanski \& Ovaskainen 2000; Ovaskainen
\emph{et al.} 2002) to estimate a distribution of time until extinction
for each landscape modification. The largest challenge in implementing
this algorithm is the space of potential modifications grows as
\(O((nm)!)\) for an \(n\) by \(m\) raster. For most actual landscapes to
which we wish to apply this method, the set of possible modifications
becomes uncomputably large, so we use simulated annealing to explore the
search space of possible modifications to estimate the modification that
maximizes the time-until extinction of simulated metapopulation dynamics
under that hypothetical modified landscape.

@@

\hypertarget{objective-1}{%
\subsection{Objective}\label{objective-1}}

@@

We need tools to generate synthetic data from a \emph{known} set of
mechanisms and parameters to test our methods for parameter inference
and forecasting on this \emph{known} system to assess the effectiveness
of these inference and forecasting methods.

\hypertarget{methods-2}{%
\subsection{Methods}\label{methods-2}}

Software is structured based on DG.jl (\textbf{cite?}) and Dispersal.jl
(\textbf{cite?}).

Uses methods from EN.jl to generate metawebs, or can use empirical
networks from Mangal.jl (\textbf{cite?}).

Framework based on (\textbf{Velland2010ConSyn?}), processes divided into
four categories: selection, dispersal, drift, speciation.

\hypertarget{results-2}{%
\subsection{Results}\label{results-2}}

\begin{figure}
\centering
\includegraphics{./figures/foodwebtraj.png}
\caption{Sample output of simulated food web dynamics from
MetacommunityDynamics.jl}
\end{figure}

\hypertarget{progress-2}{%
\subsection{Progress}\label{progress-2}}

The software as it exists is capable of simulating the biomass dynamics
of arbitrarily large food-webs using Lotka-Volterra, Holling Type-II, or
Holling Type-III functional responses. It currently has methods to
implement Guassian drift, and verious forms of dispersal via
Dispersal.jl. Also occupancy dynamics for Levins metapopulations
(\textbf{levins1967?}), and spatially explicit Hanski metapopulatoins
(\textbf{hanski2001?}). This is most of what needs to exist for the
preceding chapters.

Selection on arbitrary environmental variables in progress, as well as
traits.

@@

\hypertarget{discussion-1}{%
\section{Discussion}\label{discussion-1}}

\begin{quote}
Describing expected/anticipated contributions of the thesis. Very
important for QE. This should be at least half a page.
\end{quote}

@@
